---
title: "Data Products"
author: "J Faleiro"
date: "May 1, 2015"
output: 
    html_document:
        keep_md: true
        toc: true
        theme: united
---

# Required libraries

```{r, warning=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(googleVis, devtools, pryr)
```

# Shiny & rCharts

## rCharts example

rCharts is not on CRAN, so...

```{r}
library(devtools)
install_github('rCharts', 'ramnathv', ref = 'dev')
```

This setup is required, mainly to import CSS files that sets up width/height for the plots. It imports CSS files and JavaScript libraries from online resources. 

```{r, results='asis'}
library(rCharts)
## utility function to add required assets such as CSS and JS libraries
add_lib_assets <- function(lib, cdn = F,css=NULL) {
    assets = get_assets(get_lib(lib), cdn = cdn)
    if(!is.null(css)){assets$css=c(assets$css,css)}
    styles <- lapply(assets$css, function(style) {
        sprintf("<link rel='stylesheet' href=%s>", style)
    })

    scripts <- lapply(assets$jshead, function(script) {
        sprintf("<script type='text/javascript' src=%s></script>", script)
    })
    cat(paste(c(styles, scripts), collapse = "\n"))
}

# get assets from online repositories 
add_lib_assets("NVD3",cdn=TRUE,css="http://rawgithub.com/ramnathv/rCharts/master/inst/libraries/nvd3/css/rNVD3.css") 
add_lib_assets("Polycharts",cdn=TRUE)
```

Static plot, hover over data samples for more details

```{r, results='asis'}
names(iris) = gsub("\\.", "", names(iris))
r1<-rPlot(SepalLength ~ SepalWidth | Species, data = iris, color = 'Species', type = 'point')
r1$print("polyScatter")
```

Slightly dynamic, you can select different types (grouped, stacked) for display.

```{r nvd3mbar, results='asis',comment=NA}
hair_eye_male <- subset(as.data.frame(HairEyeColor), Sex == "Male")
n1 <- nPlot(Freq ~ Hair, group = "Eye", data = hair_eye_male, type = "multiBarChart")
n1$print("nvd3mbar")
```

## Google Vis Example

Initialization

```{r}
library(googleVis)
op <- options(gvis.plot.tag='chart')
```

Static plot:

```{r results='asis', tidy=FALSE}
CityPopularity$Mean=mean(CityPopularity$Popularity)
CC <- gvisComboChart(CityPopularity, xvar='City',
          yvar=c('Mean', 'Popularity'),
          options=list(seriesType='bars',
                       width=450, height=300,
                       title='City Popularity',
                       series='{0: {type:\"line\"}}'))
plot(CC)
```

Some animation:

```{r results='asis', tidy=FALSE}
M <- gvisMotionChart(Fruits, 'Fruit', 'Year', options=list(width=600, heigth=400))
plot(M)
options(op)
```

You can see a number of other cool plots and customizations with `library(googleViz); demo(googleViz)`

# Classes

Examples of a few classes

```{r}
class(rnorm(100))
```

```{r}
class(NA)
```

```{r}
class('foo')
```

```{r}
x <- rnorm(100)
y <- x + rnorm(100)
fit <- lm(y ~ x)
class(fit)
```

Which method is being called?

```{r}
set.seed(2)
x <- rnorm(100)
class(x)
mean(x)
```

The class of x is numeric, but there is no `mean` function for numeric objects. Se we call the default function for method mean.

```{r}
getS3method('mean', 'default')
```

S4 classes: creating a Polygon class:

```{r}
library(methods)
setClass('polygon', representation(x='numeric', y='numeric'))
```

Creating a plot `method` with `setMethod`:

```{r}
setMethod('plot', 'polygon',
          function(x, y, ...) {
              plot(x@x, x@y, type='n', ...)
              xp <- c(x@x, x@x[1])
              yp <- c(x@y, x@y[1])
              lines(xp, yp)
              })
```

And here it is:

```{r}
showMethods('plot')
```

And to get the body of this S4 method:

```{r}
getMethod('plot', 'polygon')
```

Finally, using `polygon` and `plot`:

```{r}
p <- new('polygon', x=c(1,2,3,4), y=c(1,2,3,1))
plot(p)
```

How to find out if a method is generic?

```{r}
sapply(c('mean', 'show', 'lm', 'colSums', 'dgamma', 'predict'), isGeneric)
```

```{r}
library(pryr)
c(mean=ftype(mean), show=ftype(show), lm=ftype(lm), colsums=ftype(colSums), dgamma=ftype(dgamma), predict=ftype(predict))
```
