---
title: "Reproducible Research"
author: "J Faleiro"
date: "March 19, 2015"
output: 
    html_document:
        keep_md: true
        toc: true
        theme: united
---

# Required libraries

```{r, warning=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(kernlab, UsingR, reshape, ggplot2, manipulate, dplyr, GGally, stats, rgl, car)
```

# SPAM Example

```{r}
library(kernlab)
data(spam)
```

## Data Loading and Sampling

```{r}
set.seed(3435)
trainIndicator = rbinom(nrow(spam), size=1, prob=0.5)
table(trainIndicator)
# training and testing data sets
trainSpam = spam[trainIndicator == 1, ]
testSpam = spam[trainIndicator == 0, ]
```

Some sampling:

```{r}
# number of spam/ham in training set
table(trainSpam$type)
```

Number of capital letters average per spam/ham type

## Preliminary Data Analysis

```{r}
plot(trainSpam$capitalAve ~ trainSpam$type)
```

Same, but as a logarithmic distribution, add one because we have lots of zeroes:

```{r}
plot(log10(trainSpam$capitalAve + 1) ~ trainSpam$type)
```

Pairs plot between predictors, correlation between predictors

```{r}
plot(log10(trainSpam[, 1:4] + 1))
```

## Clustering

Clustering with no transformations, almost no clusters

```{r}
hCluster = hclust(dist(t(trainSpam[, 1:57])))
plot(hCluster)
```

Clustering with transformations, several clusters

```{r}
hClusterTransformation = hclust(dist(t(log10(trainSpam[, 1:55] + 1))))
plot(hClusterTransformation)
```

## Prediction

```{r warning=FALSE, message=FALSE}
trainSpam$numType = as.numeric(trainSpam$type) - 1
costFunction = function(x, y) sum(x != (y > 0.5))
cvError = rep(NA, 55)
library(boot)
for (i in 1:55) {
    lmFormula = reformulate(names(trainSpam)[i], response="numType")
    glmFit = glm(lmFormula, family="binomial", data=trainSpam)
    cvError[i] = cv.glm(trainSpam, glmFit, costFunction, 2)$delta[2]
}
```

Which predictor has minimum cross-validated error?

```{r}
sprintf("best predictor: %s", names(trainSpam)[which.min(cvError)])
```

The best predictor is 'charDollar'.

## Measure of Uncertainty

Use best predictor found above:

```{r}
predictionModel = glm(numType ~ charDollar, family="binomial", data=trainSpam)
# get predictions on the test set
predictionTest = predict(predictionModel, testSpam)
predictedSpam = rep("nospam", dim(testSpam)[1]) # mark all as 'nospam' for now
# classify as 'spam' anything with a probability > 0.5
predictedSpam[predictionModel$fitted > 0.5] = 'spam'
# classification table
table(predictedSpam, testSpam$type)
```

Mistakes are 61 + 458, so so calculare the `errorRate`:

```{r}
errorRate <- (61 + 458) / (1346 + 458 + 61 + 449)
sprintf('error rate: %s', errorRate)
```

